### 2016-03-07 paper/icfp16-submission/semantics.tex
make genArgs mathy

\paragraph{Narrowing Types} The procedure % $\force{v}{t}{\vsu}{\tsu}$,
$$
\forcesym :: (v, t, \vsu, \tsu) \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu}
$$
defined in Figure~\ref{fig:narrow}, takes as input a value $v$, a type
$t$, and the current value and type substitutions, and refines $v$ to
have type $t$ by yielding a triple of either the same value and
substitutions, or yields the stuck state if no such refinement is
possible. In the case where $v$ is a hole, it first checks in the given
$\vsu$ to see if the hole has already been instantiated and, if so,
returns the existing instantiation.
%
\begin{figure*}[ht]
$$
\begin{array}{lcl}
%% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
\forcesym                  & ::     & (v, t, \vsu, \tsu) \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu} \\
% \force{v}{\thole}{\vsu}{\tsu}  & \defeq & \hspace{-1ex}
% \begin{cases}
%   \triple{v}{\vsu}{\tsu'} & \mbox{if } \tsu' = \unify{\{\typeof{v}, \thole\}}{\tsu} \\
%   \triple{\stuck}{\vsu}{\tsu} & \mbox{otherwise} \\

@@

\begin{figure}[ht]
$$
\begin{array}{lcll}
\gensym       & ::  & (t, \tsu) \rightarrow v \\
\gen{\thole}{\tsu}  & \defeq  & \gen{\subst{\tsu}{\thole}}{\tsu}, & \quad \text{if } \thole \in dom(\tsu) \\
\gen{\tint}{\tsu}   & \defeq  & n, & \quad \text{non-deterministic} \\
\gen{\tbool}{\tsu}  & \defeq  & b, & \quad \text{non-deterministic} \\
\gen{\ttree{t}}{\tsu}  & \defeq  & tr, & \quad \text{non-deterministic} \\
\gen{\tfun}{\tsu}   & \defeq & \efun{x}{\vhole{\thole}}, & \quad \text{\ehole and \thole are fresh} \\
\gen{\thole}{\tsu}  & \defeq & \vhole{\thole}, & \quad \text{\ehole is fresh} \\
\end{array}
$$
\caption{Generating values}
\label{fig:gen}
\end{figure}

@@

\paragraph{Modeling Semantics}
%
We resolve the non-determinism in the operational semantics
(\S~\ref{sec:semantics}) via the procedure:
%
$$
\evalN :: \triple{e}{\vsu}{\tsu} \rightarrow [\triple{v \cup \stuck}{\vsu}{\tsu}]
$$
%
Due to the non-determinism introduced by \gensym, a call
$\evalfn{\triple{e}{\vsu}{\tsu}}$ returns the \emph{list}
of possible results of the form $\triple{v \cup \stuck}{\vsu'}{\tsu'}$
such that $\steps{e}{\vsu}{\tsu}{v \cup \stuck}{\vsu'}{\tsu'}$.

\paragraph{Generating Target Arguments}
%
We address the issue of currying by defining a procedure
$\genArgs{e}$ that yields a list of holes $[v_1, \ldots, v_k]$
such that $\eapp{e}{v_1 \ldots v_k}$ \emph{does not} evaluate to
a lambda.
%
This is achieved via a loop that keeps adding holes to the
target application until evaluating the term yields a
non-lambda value.
%
% The helper functions @witness@ and @close@ are used to respectively
% apply the parameters to the target and close the result under the top-level
% let-binders, prior to invoking \hbox{@eval@.}
%
% That is, @mkApps@ creates a nested sequence of applications in
% the usual left-associative style, and @mkLets@ takes a list of
% binders and a body expression, and creates a sequence of nested
% let-binders that close the body expression.

\begin{figure}[t]
% $$
% \begin{array}{lcl}
% \genArgs{e} &\defeq& \begin{cases} \genArgs{\addArg{e}},& \mbox{if } \triple{\efun{x}{e'}}{\vsu}{\tsu},\ldots = \doeval{e} \\
%  e,& \mbox{otherwise} \\
% \end{cases}
% \\
% \addArg{e} &\defeq& \eapp{e}{\vhole{\thole}}, \qquad \mbox{where } \ehole, \thole \mbox{ are fresh }
% \end{array}
% $$

\begin{mcode}
genArgs :: $e$ -> [$v$]
genArgs $e$ = loop []
  where
  loop vs = case eval $\triple{\eapp{e}{vs}}{\emptysu}{\emptysu}$ of
             $\triple{\efun{x}{e}}{\vsu}{\tsu}$:_ -> loop ($\vhole{\thole}$ : vs)
                           -- $\ehole$, $\thole$ are fresh
             _          -> vs
\end{mcode}
\caption{Find number of arguments for target function}
\label{fig:algo-gen-args}
\end{figure}

@@

\begin{cor}{\textbf{[Witness Generation]}}
\label{thm:generation}
  If |genWitness k $e$ = $\ $ $\triple{\eapp{e}{v_1 \ldots v_n}}{\vsu}{\tsu}\ldots$|,
  then for all types $t_1 \ldots t_n$ there exist values $w_1 \ldots w_n$ such that
  $\steps{\eapp{e}{w_1 \ldots w_n}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$.
\end{cor}

\begin{figure}[t]
\centering
$$
\begin{array}{lclr}
\genWitnessN       & :: & (\mathsf{Nat} \times e) \rightarrow 2^{e} & \\
\genWitness{n}{e}  & =  & \{ \resolve{\eapp{e}{vs}}{\vsu} \mid \vsu \in \Sigma \} & \\
\quad \mbox{\textbf{where}} &    & & \\
\quad \quad vs     & =  & \genArgs{e} & (1) \\
\quad \quad res    & =  & \takefn{n}{\evalfn{\triple{\eapp{e}{vs}}{\emptysu}{\emptysu}}} & (2) \\
\quad \quad \Sigma & =  & \{ \vsu\ \mid \triple{\stuck}{\vsu}{\tsu} \in res\} & (3)
\end{array}
$$
%% \begin{mcode}
%% $\genWitnessN :: (\mathsf{Nat} \times e) \rightarrow 2^{e}$
%% $\genWitness{n}{e} = \{ \resolve{\eapp{e}{vs}}{\vsu} \mid \vsu \in \Sigma \}$
  %% where
   %% $vs     = \genArgs{e}$               -- (1)
   %% $res    = \takefn{n}{\evalfn{\triple{\eapp{e}{vs}}{\emptysu}{\emptysu}}}$  -- (2)
   %% $\Sigma = \{ \vsu\ \mid \triple{\stuck}{\vsu}{\tsu} \in res\}$   -- (3)
%% \end{mcode}
\caption{Generating Witnesses}
\label{fig:algo-gen-witness}
\end{figure}

>>>>

\paragraph{Narrowing Types} The procedure % $\force{v}{t}{\vsu}{\tsu}$,
$$
\forcesym : v \times t \times \vsu \times \tsu \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu}
$$
defined in Figure~\ref{fig:narrow}, takes as input a value $v$, a type
$t$, and the current value and type substitutions, and refines $v$ to
have type $t$ by yielding a triple of either the same value and
substitutions, or yields the stuck state if no such refinement is
possible. In the case where $v$ is a hole, it first checks in the given
$\vsu$ to see if the hole has already been instantiated and, if so,
returns the existing instantiation.
%
\begin{figure*}[ht]
$$
\begin{array}{lcl}
%% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
\forcesym                  & : & v \times t \times \vsu \times \tsu \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu} \\
% \force{v}{\thole}{\vsu}{\tsu}  & \defeq & \hspace{-1ex}
% \begin{cases}
%   \triple{v}{\vsu}{\tsu'} & \mbox{if } \tsu' = \unify{\{\typeof{v}, \thole\}}{\tsu} \\
%   \triple{\stuck}{\vsu}{\tsu} & \mbox{otherwise} \\

@@

\begin{figure}[ht]
$$
\begin{array}{lcll}
\gensym       & :   & t \times \tsu \rightarrow v \\
\gen{\thole}{\tsu}  & \defeq  & \gen{\subst{\tsu}{\thole}}{\tsu}, & \quad \text{if } \thole \in dom(\tsu) \\
\gen{\tint}{\tsu}   & \defeq  & n, & \quad \text{non-deterministic} \\
\gen{\tbool}{\tsu}  & \defeq  & b, & \quad \text{non-deterministic} \\
\gen{\ttree{t}}{\tsu}  & \defeq  & tr, & \quad \text{non-deterministic} \\
\gen{\tfun}{\tsu}   & \defeq & \efun{x}{\vhole{\thole}}, & \quad \text{\ehole and \thole are fresh} \\
\gen{\thole}{\tsu}  & \defeq & \vhole{\thole}, & \quad \text{\ehole is fresh} \\
\end{array}
$$
\caption{Generating values}
\label{fig:gen}
\end{figure}

@@

\paragraph{Modeling Semantics}
%
We resolve the non-determinism in the operational semantics
(\S~\ref{sec:semantics}) via the procedure
%
$$
\evalN : e \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu}^{*}
$$
%
Due to the non-determinism introduced by \gensym, a call
$\evalfn{e}$ returns the \emph{list}
of possible results of the form $\triple{v \cup \stuck}{\vsu}{\tsu}$
such that $\steps{e}{\emptysu}{\emptysu}{v \cup \stuck}{\vsu}{\tsu}$.

We address the issue of currying by defining a procedure \genArgs{e},
defined in Figure~\ref{fig:algo-gen-args},
that yields a list of holes $v_1, \ldots, v_n$ such that
$\eapp{e}{v_1 \ldots v_n}$ \emph{does not} evaluate to a lambda.
%
This is achieved via a loop that keeps adding holes to the
target application until evaluating the term yields a
non-lambda value.
%
% The helper functions @witness@ and @close@ are used to respectively
% apply the parameters to the target and close the result under the top-level
% let-binders, prior to invoking \hbox{@eval@.}
%
% That is, @mkApps@ creates a nested sequence of applications in
% the usual left-associative style, and @mkLets@ takes a list of
% binders and a body expression, and creates a sequence of nested
% let-binders that close the body expression.

\begin{figure}[t]
% $$
% \begin{array}{lcl}
% \genArgs{e} &\defeq& \begin{cases} \genArgs{\addArg{e}},& \mbox{if } \triple{\efun{x}{e'}}{\vsu}{\tsu},\ldots = \doeval{e} \\
%  e,& \mbox{otherwise} \\
% \end{cases}
% \\
% \addArg{e} &\defeq& \eapp{e}{\vhole{\thole}}, \qquad \mbox{where } \ehole, \thole \mbox{ are fresh }
% \end{array}
% $$
$$
\begin{array}{lcl}
\genArgsN   & : & e \rightarrow e^{*} \\
\genArgs{e} & = & \loopfn{[]} \\
\quad \mbox{\textbf{where}} &    & \\
  \quad \quad \loopfn{vs} & = & \mbox{\textbf{case }} \evalfn{\eapp{e}{vs}} \mbox{\textbf{ of}} \\
             &   & \quad \triple{\efun{x}{e}}{\vsu}{\tsu},\ldots \rightarrow \loopfn{\vhole{\thole} :: vs} \\
             &   & \qquad (\ehole, \thole \mbox{ are fresh}) \\
             &   & \quad \_ \rightarrow vs
\end{array}
$$

% \begin{mcode}
% genArgs :: $e$ -> [$v$]
% genArgs $e$ = loop []
%   where
%   loop vs = case eval $\triple{\eapp{e}{vs}}{\emptysu}{\emptysu}$ of
%              $\triple{\efun{x}{e}}{\vsu}{\tsu}$:_ -> loop ($\vhole{\thole}$ : vs)
%                            -- $\ehole$, $\thole$ are fresh
%              _          -> vs
% \end{mcode}
\caption{Finding the number of arguments for the target expression.}
\label{fig:algo-gen-args}
\end{figure}

@@

\begin{cor}{\textbf{[Witness Generation]}}
\label{thm:generation}
  If $\genWitness{k}{e} = \triple{\eapp{e}{v_1 \ldots v_n}}{\vsu}{\tsu},\ldots$,
  then for all types $t_1 \ldots t_n$ there exist values $w_1 \ldots w_n$ such that
  $\steps{\eapp{e}{w_1 \ldots w_n}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$.
\end{cor}

\begin{figure}[t]
\centering
$$
\begin{array}{lclr}
\genWitnessN       & : & \mathsf{Nat} \times e \rightarrow e^{*} & \\
\genWitness{n}{e}  & = & \{ \resolve{\eapp{e}{vs}}{\vsu} \mid \vsu \in \Sigma \} & \\
\quad \mbox{\textbf{where}} &    & & \\
\quad \quad vs     & = & \genArgs{e} & (1) \\
\quad \quad res    & = & \takefn{n}{\evalfn{\eapp{e}{vs}}} & (2) \\
\quad \quad \Sigma & = & \{ \vsu\ \mid \triple{\stuck}{\vsu}{\tsu} \in res\} & (3)
\end{array}
$$
%% \begin{mcode}
%% $\genWitnessN :: (\mathsf{Nat} \times e) \rightarrow 2^{e}$
%% $\genWitness{n}{e} = \{ \resolve{\eapp{e}{vs}}{\vsu} \mid \vsu \in \Sigma \}$
  %% where
   %% $vs     = \genArgs{e}$               -- (1)
   %% $res    = \takefn{n}{\evalfn{\triple{\eapp{e}{vs}}{\emptysu}{\emptysu}}}$  -- (2)
   %% $\Sigma = \{ \vsu\ \mid \triple{\stuck}{\vsu}{\tsu} \in res\}$   -- (3)
%% \end{mcode}
\caption{Generating witnesses.}
\label{fig:algo-gen-witness}
\end{figure}
